ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-05-17 notebooks/circles.qmd
ufff
uh

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, model) in model_dict
    # Test performance:
    _perf = CounterfactualExplanations.Models.model_evaluation(model, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"circles_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "circles_model_performance.csv"), model_performance)
model_performance
```

@@

>>>>

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, model) in model_dict
    # Test performance:
    _perf = CounterfactualExplanations.Models.model_evaluation(model, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    _perf.dataname .= "Circles"
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"circles_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "circles_model_performance.csv"), model_performance)
model_performance
```

@@

```{julia}
generator_params = DataFrame(
    Dict(
        :λ1 => λ₁,
        :λ2 => λ₂,
        :λ3 => λ₃,
        :opt => string(typeof(opt)),
        :eta => opt.eta,
        :dataname => "Circles",
    )
)
CSV.write(joinpath(params_path, "generator/circles.csv"), generator_params)
```

### 2023-05-17 notebooks/gmsc.qmd
uh

# Data:
test_size = 0.2
counterfactual_data, test_data = train_test_split(load_gmsc(); test_size=test_size)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
output_dim = length(unique(labels))
```

@@

@@

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, mod) in model_dict
    # Test performance:
    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"gmsc_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "gmsc_model_performance.csv"), model_performance)
model_performance
```

>>>>

# Data:
test_size = 0.2
counterfactual_data, test_data = train_test_split(load_gmsc(nothing); test_size=test_size)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
output_dim = length(unique(labels))
```

@@

```{julia}
params = DataFrame(
    Dict(
        :n_obs => Int.(round(n_obs/10)*10),
        :epochs => epochs,
        :batch_size => batch_size,
        :n_hidden => n_hidden,
        :n_layers => length(model_dict["MLP"].fitresult[1][1])-1,
        :activation => string(activation),
        :n_ens => n_ens,
        :lambda => string(α[3]),
        :jem_sampling_steps => jem.sampling_steps,
        :sgld_batch_size => sampler.batch_size,
        :dataname => "GMSC",
    )
)
CSV.write(joinpath(params_path, "gmsc.csv"), params)
```

@@

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, mod) in model_dict
    # Test performance:
    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    _perf.dataname .= "GMSC"
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"gmsc_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "gmsc_model_performance.csv"), model_performance)
model_performance
```

### 2023-05-17 notebooks/linearly_separable.qmd
ufff
uh

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, model) in model_dict
    # Test performance:
    _perf = CounterfactualExplanations.Models.model_evaluation(model, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"linearly_separable_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "linearly_separable_model_performance.csv"), model_performance)
model_performance
```

@@

>>>>

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, model) in model_dict
    # Test performance:
    _perf = CounterfactualExplanations.Models.model_evaluation(model, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    _perf.dataname .= "Linearly Separable"
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"linearly_separable_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "linearly_separable_model_performance.csv"), model_performance)
model_performance
```

@@

```{julia}
generator_params = DataFrame(
    Dict(
        :λ1 => λ₁,
        :λ2 => λ₂,
        :λ3 => λ₃,
        :opt => string(typeof(opt)),
        :eta => opt.eta,
        :dataname => "Linearly Separable",
    )
)
CSV.write(joinpath(params_path, "generator/linearly_separable.csv"), generator_params)
```

### 2023-05-17 notebooks/mnist.qmd
uh

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, mod) in model_dict
    # Test performance:
    test_data = load_mnist_test()
    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"mnist_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "mnist_model_performance.csv"), model_performance)
model_performance
```

>>>>

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, mod) in model_dict
    # Test performance:
    test_data = load_mnist_test()
    _perf = CounterfactualExplanations.Models.model_evaluation(mod, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    _perf.dataname .= "MNIST"
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"mnist_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "mnist_model_performance.csv"), model_performance)
model_performance
```

### 2023-05-17 notebooks/moons.qmd
ufff
uh

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, model) in model_dict
    # Test performance:
    _perf = CounterfactualExplanations.Models.model_evaluation(model, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"moons_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "moons_model_performance.csv"), model_performance)
model_performance
```

@@

>>>>

measure = Dict(
    :f1score => multiclass_f1score, 
    :acc => accuracy, 
    :precision => multiclass_precision
)
model_performance = DataFrame()
for (mod_name, model) in model_dict
    # Test performance:
    _perf = CounterfactualExplanations.Models.model_evaluation(model, test_data, measure=collect(values(measure)))
    _perf = DataFrame([[p] for p in _perf], collect(keys(measure)))
    _perf.mod_name .= mod_name
    _perf.dataname .= "Moons"
    model_performance = vcat(model_performance, _perf)
end
Serialization.serialize(joinpath(output_path,"moons_model_performance.jls"), model_performance)
CSV.write(joinpath(output_path, "moons_model_performance.csv"), model_performance)
model_performance
```

@@

```{julia}
generator_params = DataFrame(
    Dict(
        :λ1 => λ₁,
        :λ2 => λ₂,
        :λ3 => λ₃,
        :opt => string(typeof(opt)),
        :eta => opt.eta,
        :dataname => "Moons",
    )
)
CSV.write(joinpath(params_path, "generator/moons.csv"), generator_params)
```

### 2023-05-17 notebooks/tables.Rmd
uh

```{r}
files <- list.files("artifacts/params/")
dt <- lapply(files, function(x) {
    fread(file.path("artifacts/params/", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
setcolorder(
  dt, 
  c(
    "dataname", "n_obs",
    "n_hidden", "n_layers", "activation", "n_ens",
    "epochs", "batch_size",
    "jem_sampling_steps", "sgld_batch_size", "lambda"
  )
)
dt_ebm <- dt[,.(dataname, jem_sampling_steps, sgld_batch_size, lambda)]
col_names <- c(
  "Dataset",
  "SGLD Steps", "Batch Size", "$\\lambda$"
)
kbl(
  dt_ebm, caption = "EBM hyperparemeter choices for our experiments. \\label{tab:ebmparams} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex"
) %>%
  kable_styling(font_size = 8) %>%
  kable_paper(full_width = F) %>%
  save_kable("paper/contents/table_ebm_params.tex")
```

@@

>>>>

```{r}
files <- list.files("artifacts/params/")
dt <- lapply(files[grepl(".csv", files)], function(x) {
    fread(file.path("artifacts/params/", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
setcolorder(
  dt, 
  c(
    "dataname", "n_obs",
    "n_hidden", "n_layers", "activation", "n_ens",
    "epochs", "batch_size",
    "jem_sampling_steps", "sgld_batch_size", "lambda"
  )
)
dt[,dataname:=factor(dataname, levels=c("Linearly Separable", "Moons", "Circles", "MNIST", "GMSC"))]
dt <- dt[order(dataname)]
dt_ebm <- dt[,.(dataname, jem_sampling_steps, sgld_batch_size, lambda)]
col_names <- c(
  "Dataset",
  "SGLD Steps", "Batch Size", "$\\lambda$"
)
kbl(
  dt_ebm, caption = "EBM hyperparemeter choices for our experiments. \\label{tab:ebmparams} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex"
) %>%
  kable_styling(font_size = 8) %>%
  kable_paper(full_width = F) %>%
  save_kable("paper/contents/table_ebm_params.tex")
```

@@

```{r}
files <- list.files("artifacts/params/generator")
dt <- lapply(files, function(x) {
    fread(file.path("artifacts/params/generator", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt <- dt[,.(dataname,eta,λ1,λ3,λ3)]
dt[,dataname:=factor(dataname, levels=c("Linearly Separable", "Moons", "Circles", "MNIST", "GMSC"))]
dt <- dt[order(dataname)]
col_names <- c(
  "Dataset",
  "$\\eta$", "$\\lambda_1$", "$\\lambda_2$", "$\\lambda_3$"
)
kbl(
  dt, caption = "Generator hyperparameters. \\label{tab:genparams} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex"
) %>%
  kable_styling(font_size = 8) %>%
  kable_paper(full_width = F) %>%
  save_kable("paper/contents/table_gen_params.tex")
```

```{r}
files <- list.files("artifacts/results/")
dt <- lapply(files[grepl("_model_performance.csv", files)], function(x) {
    fread(file.path("artifacts/results/", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt[,dataname:=factor(dataname, levels=c("Linearly Separable", "Moons", "Circles", "MNIST", "GMSC"))]
dt <- dt[order(dataname,mod_name)]
setcolorder(
  dt, 
  c(
    "dataname", "mod_name",
    "acc", "precision", "f1score"
  )
)
col_names <- c("Dataset", "Model", "Accuracy", "Precision", "F1-Score")
kbl(
  dt, caption = "Various standard performance metrics for our different models grouped by dataset. \\label{tab:perf} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex", digits=2
) %>%
  kable_styling(font_size = 8) %>%
  kable_paper(full_width = F) %>%
  add_header_above(c(" "=2, "Performance Metrics" = 3)) %>%
  collapse_rows(columns = 1, latex_hline = "custom", valign = "top", custom_latex_hline = 1) %>%
  save_kable("paper/contents/table_perf.tex")
```

### 2023-05-17 paper/paper.tex
uh

Table~\ref{tab:params} provides an overview of all parameters related to our experiments. The \textit{GMSC} data were randomly undersampled for balancing purposes and all features were standardized. \textit{MNIST} data was also randomly undersampled for reasons outlined below. Pixel values were preprocessed to fall in the range of $[-1,1]$ and a small Gaussian noise component ($\sigma=0.03$) was added to training samples following common practices in the EBM literature.

@@

\import{contents/}{table_params.tex}
\subsection{Results}\label{app:results}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{../artifacts/results/images/mnist_eccco_all_digits.png}
  \caption{Counterfactuals for \textit{MNIST} data and our \textit{JEM Ensemble}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco}
\end{figure}

>>>>

Table~\ref{tab:params} provides an overview of all parameters related to our experiments. The \textit{GMSC} data were randomly undersampled for balancing purposes and all features were standardized. \textit{MNIST} data was also randomly undersampled for reasons outlined below. Pixel values were preprocessed to fall in the range of $[-1,1]$ and a small Gaussian noise component ($\sigma=0.03$) was added to training samples following common practice in the EBM literature. Table~\ref{tab:perf} shows standard evaluation metrics measuring the predictive performance of our different models grouped by dataset. These measures were computed over test datasets.

Table~\ref{tab:genparams} summarises our hyperparameter choices for the counterfactual generators where $\eta$ denotes the learning rate used for Stochastic Gradient Descent (SGD) and $\lambda_1$, $\lambda_2$, $\lambda_3$ represent the chosen penalty strengths (Equations~\ref{eq:general} and~\ref{eq:eccco}). Here $\lambda_1$ also refers to the chosen penalty for the distance from factual values that applies to both \textit{Wachter} and \textit{REVISE}, but not \textit{Schut} which is penalty-free. \textit{Schut} is also the only generator that uses JSMA instead of SGD for optimization.

\import{contents/}{table_params.tex}

\import{contents/}{table_perf.tex}

\import{contents/}{table_gen_params.tex}

@@

\subsection{Results}\label{app:results}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{../artifacts/results/images/mnist_eccco_all_digits.png}
  \caption{Counterfactuals for \textit{MNIST} data and our \textit{JEM Ensemble}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco}
\end{figure}
